Is structure based drug design ready for selectivity optimization?

### 2019-01-03 manuscript.tex
updating figures and small changes to grammar

\paragraph{Assessing the ability of alchemical free energy methods to predict selectivity}
Since the prediction of selectivity depends on predicting affinities to two or more targets (or relative affinities between pairs of related molecules), a spectrum of possibilities exists for the accuracy of predicted selectivity metrics.
In well-behaved kinase systems, for example, free energy calculation potency predictions have achieved mean unsigned errors of less than 1.0 kcal/mol~\citep{Wang:J.Am.Chem.Soc.:2015,Abel2017-gw}, believed to arise predominantly from systematic force field errors~\cite{schrodinger-1kcalmol-forcefield-error-paper}.
In the best-case scenario, the force field errors for two protein targets might exactly cancel out, allowing selectivity to be predicted more accurately than potency;
on the other hand, if the force field error acts like a random error between two distinct (but potentially related) protein targets, predictions of selectivity will be \emph{less accurate} than potency predictions.
Real-world systems are likely to be somewhere between these extremes, and quantifying the \emph{degree} to which error in multiple protein targets is correlated, its implications for the use of free energy calculations for prioritizing synthesis in the pursuit of selectivity, the ramifications for optimal calculation protocols, and rough guidelines governing which systems we might expect good selectivity prediction is the primary focus of this work.

@@

To demonstrate the potential impact correlation has on the forcefield error of selectivity predictions ($\sigma_{selectivity}$) using alchemical free energy techniques, we created a simple numerical model following equation~\ref{eq11}, which takes into account each of the per target forcefield errors expected from the methodology as well as the correlation in those errors. As seen in Figure ~\ref{fig:figure-1}A, if the per target forcefield errors ($\sigma_{ff,1}$ and $\sigma_{ff,2}$) are the same, $\sigma_{selectivity}$ approaches 0 as the correlation coefficient ($\rho$) approaches 1. If the error for the free energy method is not the same, $\sigma_{selectivity}$ gets smaller but approaches a non-zero value as $\rho$ approaches 1. 
To quantify the expected speedup in selectivity optimization, we modeled the change in selectivity with respect to a reference compound for a number of compounds a medicinal chemist might suggest as a normal distribution centered around 0 with a standard deviation of 1 kcal/mol (Figure~\ref{fig:figure-1}B, black curve), reflecting that most proposed modifications would not drive large changes in selectivity. Then, suppose that each compound is screened computationally with a method free energy methodology with a per target forcefield error ($\sigma_{ff}$) of 1 kcal/mol in the regime of infinite computation effort where statistical error is 0 kcal/mol. All compounds predicted to have a 1.4 kcal/mol improvement in selectivity are synthesized and experimentally tested (Figure~\ref{fig:figure-1}B, colored curves), using an experimental technique with perfect accuracy. The fold-change in the proportion of compounds that are made that have a true 1.4 kcal/mol improvement in selectivity compared to the original distribution can be calculated as a surrogate for the expected speedup. For a 1.4 kcal/mol selectivity improvement threshold (1 log unit), a correlation of 0.5 gives an expected speed up of 4.1x, which can be interpreted as 4.1x fewer compounds needing to be made before achieving a 1 log unit improvement in selectivity. This process can be extended for the even more difficult proposition of achieving a 2 log unit improvement in selectivity (Figure~\ref{fig:figure-1}C), where 200--300$\times$ speedups can be expected, depending on $\sigma_\mathrm{ff}$ for the free energy methodology.

@@

\paragraph{FEP+ calculations show accurate potency predictions for ERK2/CDK2 and larger errors for CDK2/CDK9}
The FEP+ predictions of single target potencies ($\Delta G$) showed good accuracy for the CDK2 and ERK2 dataset (Figure~\ref{fig:figure-4}, top). Replicate 1 of the calculations in shown in Figure~\ref{fig:figure-4}, with an RMSE of $0.41^{0.71}_{0.13}$ and $0.79^{1.48}_{0.09}$ kcal/mol, respectively. All of the CDK2 potencies were predicted within 1 log unit of the experimental value, while ERK2 had two outliers. The selectivity ($\Delta \Delta G_{selectivity}$) predictions show an RMSE of $0.77^{1.38}_{0.22}$ kcal/mol, with all but one of the predictions falling within 1 log unit of the experimental values (Figure~\ref{fig:figure-4}, top right panel). This was consistent across all three replicates of the calculations (Supp. Figure ~\ref{fig:sup-figure-6}). This consistency across replicates holds true at the individual ligand level as well (Supp. Figure~\ref{fig:sup-figure-8}). Despite the low RMSE for the selectivity predictions, the narrow dynamic range and high uncertainty from experiment and calculation makes it difficult to determine which compounds are more selective than others.

@@

\paragraph{Correlation of forcefield errors accelerates selectivity optimization}
To quantify the correlation coefficient ($
\rho$) of the forcefield errors in our calculations, we built a Bayesian graphical model to separate the forcefield error from the statistical error, as described in depth in the methods section. Briefly, we modeled the absolute free energy ($G$) of each ligand in each phase (complex and solvent) as in equation~\ref{eq4}. The model was chained to the FEP+ calculations by providing the $\Delta G^{calc}_{phase,ij,target}$ as observed data, as in equation~\ref{eq6}. As in equation, the experimental data was modeled as a normal distribution centered around the true free energy of binding ($\Delta G^{true}_{i,target}$) corrupted by experimental error, which is assumed to be 0.3 kcal/mol from previous work done to quantify the uncertainty in publicly available data~\citep{BROWN2009420}. The reported IC50 values from each dataset were treated as data observations (Equation~\ref{eq12}) and the $\Delta G^{true}_{i,target}$ was assigned a weak normal prior (Equation~\ref{eq13}). The correlation coefficient was calculated for each sample according to equation~\ref{eq9}. 
The correlation coefficient $\rho$ for replicate 1 of the CDK2/ERK2 calculations was quantified to be $0.49^{0.68}_{0.27}$, indicating that the errors are correlated between ERK2 and CDK2 (Figure~\ref{fig:figure-5}A, right), which was consistent with the distributions for $\rho$ in replicates 2 and 3 (Supp. Figure~\ref{fig:sup-figure-7}). The joint marginal distribution of the error ($\epsilon$) for each target is more diagonal than symmetric, which is expected for cases in which $\rho$ is 0.5 (Supp. Figure~\ref{fig:sup-figure-2}). In addition to correlation in the forcefield errors, the high per target accuracy of these calculations allow for a predicted 2-3x speed up for 1 log unit selectivity optimization, and a 20-50x speed up for 2 log unit selectivity optimization (Figure~\ref{fig:figure-5}A, right), in the regime of infinite sampling effort where statistical error is 0.

The CDK2/CDK9 calculations show strong evidence of correlation, with a correlation coefficient of $0.72^{0.83}_{0.58}$ (Figure~\ref{fig:figure-5}B, right) for replicate 1. The rest of the replicates showed strong agreement (Supp. Figure~\ref{fig:sup-figure-5}) The joint marginal distribution of errors is strongly diagonal, which is expected based on the value for $\rho$ (Figure~\ref{fig:figure-5}B, left). The high correlation in errors leads to a speed up of 2-3 for 1 log unit selectivity optimization and 30-40x for 2 log unit selectivity optimization (Figure~\ref{fig:figure-5}B, right), despite the much higher per target RMSE than the CDK2/ERK2 case.

@@

To this point, we have considered only forcefield error in quantifying the speedup free energy calculations can enable for selectivity optimization, by assuming we are in the range of infinite sampling, where the statistical error for each target is reduced to 0 kcal/mol. To begin understanding how statistical error impacts this speedup, we modified the model of speedup by additional considering the per target statistical error ($\sigma_{stat}$), which we define in Equation~\ref{eq15} such that at the baseline effort, N, $\sigma_{stat}$ is 0.2 kcal/mol. In this definition, it takes 4 times the sampling, or effort, to reduce statistical error by a factor of 2. We assume that statistical error is uncorrelated when propagating to two targets, and that $\sigma_{ff}$ is 0.9 kcal/mol for both targets~\citep{Harder2016-zn, Hauser:2018vz}. As shown in Figure~\ref{fig:figure-6}, expending effort to reduce $\sigma_{stat}$ when $\rho$ is less than 0.5 does not change the expected speedup for the 100x threshold in meaningful way, suggesting that it is not worth running calculations longer than the default protocol in this case. However, when $\rho$ is greeter than 0.5, the curves do start to separate, particularly the 1/4x, 1x and 4x effort curves. This suggests that when the correlation is high, running longer calculations can net improvements in selectivity optimization speed. Interesting, the 16x, 48x and $\infty$ effort curves do not differ greatly from the 4x effort curve, indicating that there are diminishing returns to expending more time running longer calculations.

@@

To quantify $\rho$ for the CDK2/ERK2 calculations, the default NUTS sampler with {\tt jitter+adapt\_diag} initialization, 1 000 tuning steps, and a target accept probability of 0.8 was used to draw 10 000 samples from the model. 
The CDK2/CDK9 model was sampled 20 000 times using default NUTS sampler with {\tt jitter+adapt\_diag} initialization and 3 000 tuning steps. 
\todo[inline]{JDC: I'd just rerun these both with the larger number of sampling and tuning steps. We don't want people thinking we were gaming these parameters, and we can't provide two arbitrary parameter choices without providing justification as to why.}

@@

The authors are grateful to Patrick Grinaway (ORCID: \href{http://orcid.org/0000-0002-9762-4201}{0000-0002-9762-4201}) for useful discussions about Bayesian statistics.

>>>>

\paragraph{Assessing the ability of alchemical free energy methods to predict selectivity}
Since the prediction of selectivity depends on predicting affinities to two or more targets (or relative affinities between pairs of related molecules), a spectrum of possibilities exists for the accuracy of predicted selectivity metrics.
In well-behaved kinase systems, for example, free energy calculation potency predictions have achieved mean unsigned errors of less than 1.0 kcal/mol~\citep{Wang:J.Am.Chem.Soc.:2015,Abel2017-gw}, believed to arise predominantly from systematic force field errors~\cite{Harder:J.Chem.TheoryComput.:2016}.
In the best-case scenario, the force field errors for two protein targets might exactly cancel out, allowing selectivity to be predicted more accurately than potency;
on the other hand, if the force field error acts like a random error between two distinct (but potentially related) protein targets, predictions of selectivity will be \emph{less accurate} than potency predictions.
Real-world systems are likely to be somewhere between these extremes, and quantifying the \emph{degree} to which error in multiple protein targets is correlated, its implications for the use of free energy calculations for prioritizing synthesis in the pursuit of selectivity, the ramifications for optimal calculation protocols, and rough guidelines governing which systems we might expect good selectivity prediction is the primary focus of this work.

@@

To demonstrate the potential impact correlation has on the forcefield error of selectivity predictions ($\sigma_{selectivity}$) using alchemical free energy techniques, we created a simple numerical model following Equation~\ref{eq11}, which takes into account each of the per target forcefield errors expected from the methodology as well as the correlation in those errors. As seen in Figure ~\ref{fig:figure-1}A, if the per target forcefield errors ($\sigma_{ff,1}$ and $\sigma_{ff,2}$) are the same, $\sigma_{selectivity}$ approaches 0 as the correlation coefficient ($\rho$) approaches 1. If the error for the free energy method is not the same, $\sigma_{selectivity}$ gets smaller but approaches a non-zero value as $\rho$ approaches 1. 
To quantify the expected speedup in selectivity optimization, we modeled the change in selectivity with respect to a reference compound for a number of compounds a medicinal chemist might suggest as a normal distribution centered around 0 with a standard deviation of 1 kcal/mol (Figure~\ref{fig:figure-1}B, black curve), reflecting that most proposed modifications would not drive large changes in selectivity. Then, suppose that each compound is screened computationally with a free energy methodology with a per target forcefield error ($\sigma_{ff}$) of 1 kcal/mol in the regime of infinite computational effort where statistical error is 0 kcal/mol. All compounds predicted to have a 1.4 kcal/mol improvement in selectivity are synthesized and experimentally tested (Figure~\ref{fig:figure-1}B, colored curves), using an experimental technique with perfect accuracy. The fold-change in the proportion of compounds that are made that have a true 1.4 kcal/mol improvement in selectivity compared to the original distribution can be calculated as a surrogate for the expected speedup. For a 1.4 kcal/mol selectivity improvement threshold (1 log unit), a correlation of 0.5 gives an expected speed up of 4.1x, which can be interpreted as need to make 4.1x fewer compounds to achieve a 1 log unit improvement in selectivity. This process can be extended for the even more difficult proposition of achieving a 2 log unit improvement in selectivity (Figure~\ref{fig:figure-1}C), where 200--300$\times$ speedups can be expected, depending on $\sigma_\mathrm{ff}$ for the free energy methodology.

@@

\paragraph{FEP+ calculations show accurate potency predictions for ERK2/CDK2 and larger errors for CDK2/CDK9}
The FEP+ predictions of single target potencies ($\Delta G$) showed good accuracy for the CDK2 and ERK2 dataset (Figure~\ref{fig:figure-4}, top). Replicate 1 of the calculations is shown in Figure~\ref{fig:figure-4}, with an RMSE of $0.41^{0.71}_{0.13}$ and $0.79^{1.48}_{0.09}$ kcal/mol, respectively. All of the CDK2 potencies were predicted within 1 log unit of the experimental value, while ERK2 had two outliers. The selectivity ($\Delta \Delta G_{selectivity}$) predictions show an RMSE of $0.77^{1.38}_{0.22}$ kcal/mol, with all but one of the predictions falling within 1 log unit of the experimental values (Figure~\ref{fig:figure-4}, top right panel). This was consistent across all three replicates of the calculations (Supp. Figure ~\ref{fig:sup-figure-6}). This consistency across replicates holds true at the individual ligand level as well (Supp. Figure~\ref{fig:sup-figure-8}). Despite the low RMSE for the selectivity predictions, the narrow dynamic range and high uncertainty from experiment and calculation makes it difficult to determine which compounds are more selective than others.

@@

\paragraph{Correlation of forcefield errors accelerates selectivity optimization}
To quantify the correlation coefficient ($
\rho$) of the forcefield errors in our calculations, we built a Bayesian graphical model to separate the forcefield error from the statistical error, as described in depth in the methods section. Briefly, we modeled the absolute free energy ($G$) of each ligand in each phase (complex and solvent) as in equation~\ref{eq4}. The model was chained to the FEP+ calculations by providing the $\Delta G^{calc}_{phase,ij,target}$ as observed data, as in equation~\ref{eq6}. As in equation, the experimental data was modeled as a normal distribution centered around the true free energy of binding ($\Delta G^{true}_{i,target}$) corrupted by experimental error, which is assumed to be 0.3 kcal/mol from previous work done to quantify the uncertainty in publicly available data~\citep{BROWN2009420}. The reported IC50 values from each dataset were treated as data observations (Equation~\ref{eq12}) and the $\Delta G^{true}_{i,target}$ was assigned a weak normal prior (Equation~\ref{eq13}). The correlation coefficient was calculated for each sample according to equation~\ref{eq9}. 
The correlation coefficient $\rho$ for replicate 1 of the CDK2/ERK2 calculations was quantified to be $0.49^{0.68}_{0.27}$, indicating that the errors are correlated between ERK2 and CDK2 (Figure~\ref{fig:figure-5}A, right), which was consistent with the distributions for $\rho$ in replicates 2 and 3 (Supp. Figure~\ref{fig:sup-figure-7}). The joint marginal distribution of the error ($\epsilon$) for each target is more diagonal than symmetric, which is expected for cases in which $\rho$ is 0.5 (Supp. Figure~\ref{fig:sup-figure-2}). In addition to correlation in the forcefield errors, the high per target accuracy of these calculations allow for a predicted 2-3x speed up for 1 log unit selectivity optimization, and a 20-50x speed up for 2 log unit selectivity optimization (Figure~\ref{fig:figure-5}A, right), in the regime of infinite sampling effort where there is no statistical error.

The CDK2/CDK9 calculations show strong evidence of correlation, with a correlation coefficient of $0.72^{0.83}_{0.58}$ (Figure~\ref{fig:figure-5}B, right) for replicate 1. The rest of the replicates showed strong agreement (Supp. Figure~\ref{fig:sup-figure-5}). The joint marginal distribution of errors is strongly diagonal, which is expected based on the value for $\rho$ (Figure~\ref{fig:figure-5}B, left). The high correlation in errors leads to a speed up of 2-3 for 1 log unit selectivity optimization and 30-40x for 2 log unit selectivity optimization (Figure~\ref{fig:figure-5}B, right), despite the much higher per target RMSE than the CDK2/ERK2 case.

@@

To this point, we have considered only forcefield error in quantifying the speedup free energy calculations can enable for selectivity optimization, by assuming we are in the range of infinite sampling, where the statistical error for each target is reduced to 0 kcal/mol. To begin understanding how statistical error impacts this speedup, we modified the model of speedup by additionally considering the per target statistical error ($\sigma_{stat}$), which we define in Equation~\ref{eq15} such that at the baseline effort, N, $\sigma_{stat}$ is 0.2 kcal/mol. In this definition, it takes 4 times the sampling, or effort, to reduce statistical error by a factor of 2. We assume that statistical error is uncorrelated when propagating to two targets, and that $\sigma_{ff}$ is 0.9 kcal/mol for both targets~\citep{Harder2016-zn, Hauser:2018vz}. As shown in Figure~\ref{fig:figure-6}, expending effort to reduce $\sigma_{stat}$ when $\rho$ is less than 0.5 does not change the expected speedup for the 100x threshold in meaningful way, suggesting that it is not worth running calculations longer than the default protocol in this case. However, when $\rho$ is greeter than 0.5, the curves do start to separate, particularly the 1/4x, 1x and 4x effort curves. This suggests that when the correlation is high, running longer calculations can net improvements in selectivity optimization speed. Interestingly, the 16x, 48x and $\infty$ effort curves do not differ greatly from the 4x effort curve, indicating that there are diminishing returns to running longer calculations.

@@

To quantify $\rho$ from these calculations, the default NUTS sampler with {\tt jitter+adapt\_diag} initialization, 3 000 tuning steps, and the default target accept probability was used to draw 20 000 samples from the model. 
\todo[inline]{JDC: I'd just rerun these both with the larger number of sampling and tuning steps. We don't want people thinking we were gaming these parameters, and we can't provide two arbitrary parameter choices without providing justification as to why.}
\todo[inline]{SKA: done! The results are basically the same. I will update the figures shortly. The text above has been modified to reflect what was done.}

@@

The authors are grateful to Patrick Grinaway (ORCID: \href{http://orcid.org/0000-0002-9762-4201}{0000-0002-9762-4201}) for useful discussions about Bayesian statistics and Mehtap Işık (ORCID: \href{https://orcid.org/0000-0002-6789-952X}{0000-0002-6789-952X}) for useful discussion about kinase inhibitor protonation states. SKA is grateful to  Haoyu S. Yu, Wei Chen, and Dmitry Lupyan for advice on running FEP+ calculations.
